---
title: 'DSPA2: Data Science and Predictive Analytics (UMich HS650)'
author: "<h3>SOCR/MIDAS (Yuan-Yu Lin)</h3>"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_document: null
  theme: spacelab
  pdf_document: default
tags:
- DSPA
- SOCR
- MIDAS
- Big Data
- Predictive Analytics
subtitle: <h2><u>AI analytics model for heart failure predictions</u></h2>
highlight: tango
includes:
  before_body: SOCR_header.html
toc: true
number_sections: true
toc_depth: 3
toc_float:
  collapsed: false
smooth_scroll: true
code_folding: show
self_contained: true
---

```{r}
library(plotly)
library(ggplot2)
library(gmodels)
library(psych)
```

## import CSV

### The data from kaggle.com

#### <https://www.kaggle.com/datasets/fedesoriano/heart-failure-prediction>

```{r}
raw_data <- read.csv("heart_failure_prediction_dataset.csv")
raw_data

```

## overview the dataset

```{r}
#filter zero in the column of cholesterol

raw_data <- raw_data[raw_data$Cholesterol !=0, ]

Age <- raw_data$Age
RestingBP <- raw_data$RestingBP
Cholesterol <- raw_data$Cholesterol
MaxHR <- raw_data$MaxHR

head(raw_data)
summary(raw_data)
str(raw_data)
```

## Visualization of dataset (plotly)

```{r}
# Age 
summary(raw_data$Age)
age_density<- density(raw_data$Age)
plot_ly(x=age_density$x, y=age_density$y, type = "scatter", mode = "line", fill = "tozeroy", name = "Density") %>%
  layout(title='Density of (%)Age', xaxis = list (title = 'Age'), yaxis = list (title = 'Density'))

## The distribution of Age 
mean_age <- mean(raw_data$Age)
sd_age <-sd(raw_data$Age)
rand1000_age <- rnorm(10000, mean=mean_age, sd=sd_age)
age_density_distribution <- density(rand1000_age, bw =1)
age_histogram <-plot_ly(x=~raw_data$Age, type="histogram", histnorm = "probability", name = "Age Histogram") %>% add_trace(x=~age_density_distribution$x, y=~2.5*age_density_distribution$y,type = "scatter", mode = "lines", opacity=0.1, fill = "tozeroy", name = "Normal Density") 

  #add label and  the 68−95−99.7 line.
age_histogram<-age_histogram %>% layout(title='Age Histogram & Model Distribution', xaxis = list(title = "Age"), yaxis = list(title = "relative frequency/density"), legend = list(orientation = 'h'), shapes = list(list(type = "line", x0 = mean_age, x1 = mean_age, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "black", dash="dash")),list(type = "line", x0 = mean_age - sd_age, x1 = mean_age - sd_age, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "red", dash="dot")), list(type = "line", x0 = mean_age + sd_age, x1 = mean_age + sd_age, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "red", dash="dot")), list(type = "line", x0 = mean_age - 2*sd_age, x1 = mean_age - 2*sd_age, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "blue", dash="dot")),
      list(type = "line", x0 = mean_age + 2*sd_age, x1 = mean_age + 2*sd_age, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "blue", dash="dot")),
      list(type = "line", x0 = mean_age - 3*sd_age, x1 = mean_age - 3*sd_age, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "green", dash="dot")),
      list(type = "line", x0 = mean_age + 3*sd_age, x1 = mean_age + 3*sd_age, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "green", dash="dot"))))
age_histogram


```

```{r}
#Rest BP
summary(raw_data$RestingBP)
restBP_density<- density(raw_data$RestingBP)
plot_ly(x=restBP_density$x, y=restBP_density$y, type = "scatter", mode = "line", fill = "tozeroy", name = "Density") %>%
  layout(title='Density of (%)Rest BP', xaxis = list (title = 'Resting Blood Pressure'), yaxis = list (title = 'Density'))


## The distribution of resting BP
mean_restBP <- mean(raw_data$RestingBP)
sd_restBP <-sd(raw_data$RestingBP)
rand1000_restBP <- rnorm(10000, mean=mean_restBP, sd=sd_restBP)
restBP_density_distribution <- density(rand1000_restBP, bw =1)
restBP_histogram <-plot_ly(x=~raw_data$RestingBP, type="histogram", histnorm = "probability", name = "Resting Blood pressure Histogram") %>%
  add_trace(x=~restBP_density_distribution$x, y=~6*restBP_density_distribution$y,type = "scatter", mode = "lines", opacity=0.1, fill = "tozeroy", name = "Normal Density") 


  #add label and  the 68−95−99.7 line.
restBP_histogram <-restBP_histogram %>% layout(title='resting Blood pressure Histogram &  Distribution', xaxis = list(title = "resting blood pressure"), yaxis = list(title = "relative frequency/density"), legend = list(orientation = 'h'), shapes = list(list(type = "line", x0 = mean_restBP, x1 = mean_restBP, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "black", dash="dash")),list(type = "line", x0 = mean_restBP - sd_restBP, x1 = mean_restBP - sd_restBP, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "red", dash="dot")), list(type = "line", x0 = mean_restBP + sd_restBP, x1 = mean_restBP + sd_restBP, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "red", dash="dot")), list(type = "line", x0 = mean_restBP - 2*sd_restBP, x1 = mean_restBP - 2*sd_restBP, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "blue", dash="dot")),
      list(type = "line", x0 = mean_restBP + 2*sd_restBP, x1 = mean_restBP + 2*sd_restBP, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "blue", dash="dot")),
      list(type = "line", x0 = mean_restBP - 3*sd_restBP, x1 = mean_restBP - 3*sd_restBP, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "green", dash="dot")),
      list(type = "line", x0 = mean_restBP + 3*sd_restBP, x1 = mean_restBP + 3*sd_restBP, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "green", dash="dot"))))
restBP_histogram

```

```{r}
## Function plot_histogram for Probability density
plot_histogram <- function(data, density_distribution, data_column ,mean_, sd_, graph_title, y_axis_weight=1) {
  plot_ly(x = ~data, type = "histogram", histnorm = "probability") %>%
    add_trace(
      x = ~density_distribution$x,
      y = ~ y_axis_weight*density_distribution$y,
      type = "scatter",
      mode = "lines",
      opacity = 0.1,
      fill = "tozeroy",
      name = "Normal Density"
    ) %>%
    layout(
      title = graph_title,
      xaxis = list(title = data_column ),
      yaxis = list(title = "Relative Frequency / Density"),
      legend = list(orientation = 'h'),
      shapes = list(
        list(type = "line", x0 = mean_, x1 = mean_, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "black", dash = "dash")),
        list(type = "line", x0 = mean_ - sd_, x1 = mean_ - sd_, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "red", dash = "dot")),
        list(type = "line", x0 = mean_ + sd_, x1 = mean_ + sd_, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "red", dash = "dot")),
        list(type = "line", x0 = mean_ - 2 * sd_, x1 = mean_ - 2 * sd_, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "blue", dash = "dot")),
        list(type = "line", x0 = mean_ + 2 * sd_, x1 = mean_ + 2 * sd_, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "blue", dash = "dot")),
        list(type = "line", x0 = mean_ - 3 * sd_, x1 = mean_ - 3 * sd_, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "green", dash = "dot")),
        list(type = "line", x0 = mean_ + 3 * sd_, x1 = mean_ + 3 * sd_, y0 = 0, y1 = 1, xref = "x", yref = "paper", line = list(color = "green", dash = "dot"))
      )
    )
}

raw_data$Cholesterol

#cholesterol 
summary(raw_data$Cholesterol,na.rm = T)

cholesterol_density <- density(raw_data$Cholesterol, na.rm = T)
cholesterol_density
plot_ly(x=cholesterol_density$x, y=cholesterol_density$y, type = "scatter", mode = "line", fill = "tozeroy", name = "Density") %>%
  layout(title='Density of (%)Chelesterol', xaxis = list (title = 'cholesterol'), yaxis = list (title = 'Density'))

## The distribution of cholesterol
mean_cholesterol <- mean(raw_data$Cholesterol)
sd_cholesterol <-sd(raw_data$Cholesterol)
rand1000_colesterol <- rnorm(10000, mean=mean_cholesterol, sd=sd_cholesterol)
cholesterol_density_distribution <- density(rand1000_colesterol, bw =1)
cholesterol_histogram <- plot_histogram(raw_data$Cholesterol, cholesterol_density_distribution, "Cholesterol", mean_cholesterol, sd_cholesterol, "Cholesterol Histogram & Distribution", 8)

cholesterol_histogram
```

```{r}
#Max_HR
summary(raw_data$MaxHR)
max_HR <- density(raw_data$MaxHR)
plot_ly(x=max_HR$x, y=max_HR$y, type = "scatter", mode = "line", fill = "tozeroy", name = "Density") %>%
  layout(title='Density of max HR', xaxis = list (title = 'Max Heart Rate'), yaxis = list (title = 'Density'))

mean_maxHR <- mean(raw_data$MaxHR)
sd_maxHR <-sd(raw_data$MaxHR)
rand1000_MaxHR <- rnorm(10000, mean=mean_maxHR, sd=sd_maxHR)
maxHR_density_distribution <- density(rand1000_MaxHR, bw =1)
maxHR_histogram <- plot_histogram(raw_data$MaxHR, maxHR_density_distribution, "MaxHR", mean_maxHR, sd_maxHR, "Max Heart Rate Histogram & Distribution", 6)

maxHR_histogram
```

## Exploring Relationships Between Variables

```{r}
# age and Resting BP
scatter_plot_age_resting_BP <- plot_ly(x = ~raw_data$Age, y = ~raw_data$RestingBP, type = "scatter", 
        mode = "markers") %>% layout(title='Age vs. Resting BP', 
                                         xaxis = list (title = 'Age'), yaxis = list (title = 'Resting BP'))
scatter_plot_age_resting_BP


# age and Cholesterol 
scatter_plot_age_cholesterol <- plot_ly(x = ~raw_data$Age, y = ~raw_data$Cholesterol, type = "scatter", 
        mode = "markers") %>% layout(title='Age vs. Cholesterol', 
                                         xaxis = list (title = 'Age'), yaxis = list (title = 'Cholesterol'))
scatter_plot_age_cholesterol

# Age and MaxHR
scatter_plot_age_maxHR <- plot_ly(x = ~raw_data$Age, y = ~raw_data$MaxHR, type = "scatter", 
        mode = "markers") %>% layout(title='Age vs. Max HR', 
                                         xaxis = list (title = 'Age'), yaxis = list (title = 'Max HR'))
scatter_plot_age_maxHR


# RestingBP and Cholesterol
scatter_plot_restingBP_cholesterol <- plot_ly(x = ~raw_data$RestingBP, y = ~raw_data$Cholesterol, type = "scatter", 
        mode = "markers") %>% layout(title='resting BP vs. cholesterol', 
                                         xaxis = list (title = 'resting BP'), yaxis = list (title = 'cholesterol'))
scatter_plot_restingBP_cholesterol 

# Resting BP and Max HR
scatter_plot_restingBP_maxHR <- plot_ly(x = ~raw_data$RestingBP, y = ~raw_data$MaxHR, type = "scatter", 
        mode = "markers") %>% layout(title='resting BP vs. Max HR', 
                                         xaxis = list (title = 'Resting BP'), yaxis = list (title = 'Max HR'))
scatter_plot_restingBP_maxHR


# Cholesterol MaxHR
scatter_plot_cholesterol_maxHR <- plot_ly(x = ~raw_data$Cholesterol, y = ~raw_data$MaxHR, type = "scatter", 
        mode = "markers") %>% layout(title='cholesterol vs. Max HR', 
                                         xaxis = list (title = 'cholesterol'), yaxis = list (title = 'Max HR'))
scatter_plot_cholesterol_maxHR


```
Relationship between distinct features
```{r}
feature_relationship <- function(x_, y_, z1_, z2_, title_name, x_label, y_label, z_label, z1_label, z2_label){
  plot_ly() %>% 
    add_trace(x = x_, y = y_, z = z1_, type = 'scatter3d', mode = 'markers', marker = list(color = z1_, colorscale = 'Viridis', opacity = 0.7,size = ~z1_ / 25), name = z1_label) %>%
    add_trace(x = x_, y = y_, z = z2_, type = 'scatter3d', mode = 'markers', marker = list(color = z2_, colorscale = 'Cividis', opacity = 0.7,size = ~z1_ / 25), name = z2_label) %>% 
    layout(title = title_name, scene = list(xaxis = list(title = x_label), yaxis = list(title = y_label), zaxis = list(title = z_label)))}


###### ========================
#  Between Age and Resting Blood Pressure 
  # by Cholesterol Level
plot_ly(raw_data) %>%
    add_markers(x = ~Age, y = ~RestingBP, type = "scatter",  color = ~Cholesterol)%>% layout(title = "Relationship Between Age and Resting Blood Pressure by Cholesterol Level" )
  # by Max HR Level
plot_ly(raw_data) %>%
    add_markers(x = ~Age, y = ~RestingBP, type = "scatter",  color = ~MaxHR)%>% layout(title = "Relationship Between Age and Resting Blood Pressure by Max Heart Rate Level" )
  # by  Cholesterol Level and Max HR Level
Age_restingBP_by_Cholesterol_MaxHR <-feature_relationship(Age,RestingBP, Cholesterol,MaxHR, title_name = "Relationship Between Age and Resting Blood Pressure", "Age", "Resting BP", "Cholesterol /Max HR", "Cholesterol", "MaxHR" )
Age_restingBP_by_Cholesterol_MaxHR


###### ========================
# Between Age and Cholesterol 
  # by Resting BP Level
plot_ly(raw_data) %>%
    add_markers(x = ~Age, y = ~Cholesterol, type = "scatter",  color = ~RestingBP)%>% layout(title = "Relationship Between Age and Cholesterol by Resting BP Level" )
  # by Max HR Level
plot_ly(raw_data) %>%
    add_markers(x = ~Age, y = ~Cholesterol, type = "scatter",  color = ~MaxHR)%>% layout(title = "Relationship Between Age and Cholesterol by Max Heart Rate Level" )
  # by Max HR Level and  resting BP
Age_choleesterol_by_restingBP_MaxHR <-feature_relationship(Age,Cholesterol,RestingBP, MaxHR, title_name = "Relationship Between Age and Cholesterol", "Age", "Cholesterol", "Resting BP /Max HR","RestingBP","MaxHR" )
Age_choleesterol_by_restingBP_MaxHR 


###### ========================
# Between Age and Max HR 
  # by Resting BP Level
plot_ly(raw_data) %>%
    add_markers(x = ~Age, y = ~MaxHR, type = "scatter",  color = ~RestingBP)%>% layout(title = "Relationship Between Age and MaxHR by Resting BP Level" )
  # by Cholesterol level
plot_ly(raw_data) %>%
    add_markers(x = ~Age, y = ~MaxHR, type = "scatter",  color = ~Cholesterol)%>% layout(title = "Relationship Between Age and MaxHR by Cholesterol Level" )
  # by Resting BP Level and Cholesterol level
Age_MaxHR_by_restingBP_Cholesterol <-feature_relationship(Age,MaxHR,RestingBP,Cholesterol, title_name = "Relationship Between Age and MaxHR", "Age", "MaxHR", "Resting BP /Cholesterol", "RestingBP", "Cholesterol" )
Age_MaxHR_by_restingBP_Cholesterol


###### ========================
# Between Resting BP and MaxHR 
  # by age Level
plot_ly(raw_data) %>%
    add_markers(x = ~RestingBP, y = ~MaxHR, type = "scatter",  color = ~Age)%>% layout(title = "Relationship Between Resting BP and MaxHR by Age Level" )
  # by Cholesterol Level
plot_ly(raw_data) %>%
    add_markers(x = ~RestingBP, y = ~MaxHR, type = "scatter",  color = ~Age)%>% layout(title = "Relationship Between Resting BP and MaxHR by Age Level" )
  # by age and Cholesterol
RestingBP_MaxHR_by_Age_Cholesterol <-feature_relationship(RestingBP, MaxHR, Age,Cholesterol, title_name = "Relationship Between Resting BP and MaxHR", "RestingBP", "MaxHR", "Age/Cholesterol", "Age", "Cholesterol" )
RestingBP_MaxHR_by_Age_Cholesterol


###### ========================
# Between Resting BP and Cholesterol 
  # by MaxHR Level
plot_ly(raw_data) %>%
    add_markers(x = ~RestingBP, y = ~Cholesterol, type = "scatter",  color = ~MaxHR)%>% layout(title = "Relationship Between Resting BP and Cholesterol by MaxHR Level" )
  # by Age Level
plot_ly(raw_data) %>%
    add_markers(x = ~RestingBP, y = ~Cholesterol, type = "scatter",  color = ~Age)%>% layout(title = "Relationship Between Resting BP and Cholesterol by Age Level" )
  # by MaxHR and Age
RestingBP_Cholesterol_by_MaxHR_Age <-feature_relationship(RestingBP, Cholesterol, MaxHR, Age, title_name = "Relationship Between Resting BP and Cholesterol", "RestingBP", "Cholesterol", "MaxHR/Age", "MaxHR", "Age" )
RestingBP_Cholesterol_by_MaxHR_Age


# Between Cholesterol and MaxHR 
  # by age Level
plot_ly(raw_data) %>%
    add_markers(x = ~Cholesterol, y = ~MaxHR, type = "scatter",  color = ~Age)%>% layout(title = "Relationship Between Cholesterol and MaxHR by Age Level" )
  # by resting BP Level
plot_ly(raw_data) %>%
    add_markers(x = ~Cholesterol, y = ~MaxHR, type = "scatter",  color = ~RestingBP)%>% layout(title = "Relationship Between Cholesterol and MaxHR by Resting BP Level" )
  # By age and resting BP
RestingBP_Cholesterol_by_MaxHR_Age <-feature_relationship(Cholesterol, MaxHR,   RestingBP, Age, title_name = "Relationship Between Cholesterol and MaxHR ", "Cholesterol", "MaxHR", "RestingBP/Age ", "RestingBP", "Age" )
RestingBP_Cholesterol_by_MaxHR_Age


```


## correlation matrix between age, rest BP, cholesterol, max HR

```{r}
list_feature <- c("Age", "RestingBP", "Cholesterol", "MaxHR")

cor(raw_data[list_feature])


```

## Visualizing relationships

```{r}
clean_data_feature <- raw_data[, c("Age", "RestingBP", "Cholesterol", "MaxHR")] 
feature_visualization <- plot_ly(raw_data) %>%
  add_trace(
    type = 'splom',
    dimensions = list(
      list(label = 'Age', values = ~Age),
      list(label = 'Resting BP', values = ~RestingBP),
      list(label = 'Cholesterol', values = ~Cholesterol),
      list(label = 'MaxHR', values = ~MaxHR)  # <- fixed typo: value -> values
    ),
    marker = list(  # Set all points to blue
      size = 7,
      line = list(width = 1, color = 'rgb(230,230,230)')
    )
  ) %>%
  layout(
    title = 'Feature Relationship',
    plot_bgcolor = 'rgba(240,240,240,0.95)'
  )
feature_visualization

```

## scatter plot matrix

```{r}
pairs.panels(raw_data[, c("Age", "RestingBP", "Cholesterol", "MaxHR")])
```

# Relationship between various feature in heatmap

```{r}
filtered_no_heart<-raw_data[, c("Age", "RestingBP", "Cholesterol", "MaxHR")]
cor_matrix <- cor(filtered_no_heart)
Four_feature_heat_map<- plot_ly(x =~colnames(cor_matrix), y = ~rownames(cor_matrix), z = cor_matrix, type = "heatmap")%>%
  layout(title="correlation heatmap of four features ", xaxis=list(title="Fearture"), yaxis=list(title="Feature"))
Four_feature_heat_map



```




## Training a model on the data

## 75% data as training data and 25% data as test data

```{r}


filter_data <-raw_data[, c("HeartDisease", "Age", "RestingBP", "Cholesterol", "MaxHR")] 
n <- nrow(filter_data)
train_indices <- sample(seq_len(n), size = 0.75 * n)
training_data <- filter_data[train_indices, ]
test_data <- filter_data[-train_indices, ]


head(training_data)
summary(training_data)
str(training_data)

head(test_data)
summary(test_data)
str(test_data)

```

## mathematical modeling - logistic regression

```{r}
#"Age", "RestingBP", "Cholesterol", "MaxHR" as the independent variable
# Heart Disease as the dependent variable

Y <- training_data$HeartDisease
Y
X <- cbind(rep(1, n=nrow(training_data)), training_data$Age, training_data$RestingBP, training_data$Cholesterol, training_data$ MaxHR)
beta <- solve(t(X) %*% X) %*% t(X) %*% Y
beta

predict_model <- glm(HeartDisease~Age+RestingBP+Cholesterol+MaxHR, data=training_data, family = binomial)
summary(predict_model)

predicted_probs <- predict(predict_model, type = "response")
predicted_classes <- ifelse(predicted_probs > 0.8, 1, 0)

```
